---
title: "initial_exploration"
author: "D Perret"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      prompt = FALSE,
                      error = TRUE,
                      message = FALSE,
                      cache = FALSE,
                      cache.lazy = FALSE,
                      eval = TRUE,
                      eval.after = "fig.cap")

library(tidyverse)
library(rFIA)
library(ggplot2)
library(ggExtra)
library(rgdal)
library(sp)
library(ggsci)
library(raster)
library(splines)
library(lme4)
library(brms)
library(patchwork)
library(performance)
library(ggeffects)
select <- dplyr::select

# setting my preferred ggplot2 theme
theme_set(theme_bw())
theme_update(text = element_text(size=16, color = "black"),
             panel.grid.major = element_blank(),
             panel.grid.minor = element_blank(),
             strip.background = element_blank(),
             panel.border=element_rect(size=1.5))

# colors to be used in all map figures
mapcolor <- "wheat3"
linecolor <- "gray40"
regioncolors <- c("Cascade Mixed Forest" = "#009E73",
                  "Northern Rocky Mountain Forest-Steppe" = "#56B4E9",
                  "Middle Rocky Mountain Steppe" = "#E69F00",
                  "Southern Rocky Mountain Steppe" = "#CC79A7",
                  "zIntermountain Semi-Desert" = "gray20",
                  "AZ-NM Mountains" = "#D55E00")

# sourcing estimation base code
source("code/growMort_rewrite_METRIC.R")

```

Loading in FIA data

```{r}
fia <- readFIA(dir = "data/FIA/",common = T)

## some data curation

fia$PLOT <- fia$PLOT %>% 
  mutate(pltID = paste(UNITCD,STATECD,COUNTYCD,PLOT,sep="_"),
         PLT_CN = CN,
         #ECOSUBCD = trimws(ECOSUBCD),
         state_key = case_when(STATECD == 37 ~ "NC",                # ADD IN THE OTHER STATECDS LATER
                               STATECD == 45 ~ "SC",
                               STATECD == 51 ~ "VA",
                               STATECD == 47 ~ "TN",
                               STATECD == 21 ~ "KY",
                               STATECD == 54 ~ "WV",
                               STATECD == 13 ~ "GA",
                               STATECD == 1 ~ "AL")) %>% 
  group_by(pltID) %>% 
  mutate(most.recent = ifelse(MEASYEAR==max(MEASYEAR),
                              "yes","no")) %>% 
  ungroup()

fia$TREE <- fia$TREE %>% 
  left_join(fia$PLOT %>% 
              select(PLT_CN,most.recent,state_key),
            by="PLT_CN") %>% 
  mutate(TRE_CN = CN,
         #DIA = DIA*2.54,
         #PREVDIA = PREVDIA*2.54,
         agent_key = case_when(STATUSCD==2 & AGENTCD %in% c(00,70) ~ "unknown1",
                               STATUSCD==2 & AGENTCD == 10 ~ "insect",
                               STATUSCD==2 & AGENTCD == 20 ~ "disease",
                               STATUSCD==2 & AGENTCD == 30 ~ "fire",
                               STATUSCD==2 & AGENTCD == 40 ~ "animal",
                               STATUSCD==2 & AGENTCD == 50 ~ "weather",
                               STATUSCD==2 & AGENTCD == 60 ~ "competition",
                               STATUSCD==2 & AGENTCD == 80 ~ "land use",
                               STATUSCD==2 & is.na(AGENTCD) & 
                                 (PREV_STATUS_CD==1 | 
                                    is.na(PREV_STATUS_CD)) ~ "unknown2"),
         insect.damage = case_when(DAMAGE_AGENT_CD1 >= 10000 &
                                     DAMAGE_AGENT_CD1 < 19000 ~ 1,
                                   DAMAGE_AGENT_CD2 >= 10000 &
                                     DAMAGE_AGENT_CD2 < 19000 ~ 1,
                                   DAMAGE_AGENT_CD3 >= 10000 &
                                     DAMAGE_AGENT_CD3 < 19000 ~ 1,
                                   TRUE ~ 0),
         disease.damage = case_when(DAMAGE_AGENT_CD1 >= 20000 &
                                      DAMAGE_AGENT_CD1 < 30000 ~ 1,
                                    DAMAGE_AGENT_CD2 >= 20000 &
                                      DAMAGE_AGENT_CD2 < 30000 ~ 1,
                                    DAMAGE_AGENT_CD3 >= 20000 &
                                      DAMAGE_AGENT_CD3 < 30000 ~ 1,
                                    TRUE ~ 0),
         other.damage = case_when(DAMAGE_AGENT_CD1 > 30000 ~ 1,
                                  DAMAGE_AGENT_CD2 > 30000 ~ 1,
                                  DAMAGE_AGENT_CD3 > 30000 ~ 1,
                                  TRUE ~ 0)) %>% 
  left_join(.,
            fia$TREE %>% 
              select(PREV_TRE_CN, SPCD) %>% 
              rename(LATER_SPCD=SPCD),
            by=c("TRE_CN"="PREV_TRE_CN")) %>% 
  mutate(SPCD = case_when(SPCD!=LATER_SPCD & !is.na(LATER_SPCD) ~ LATER_SPCD,
                          is.na(LATER_SPCD) ~ SPCD,
                          TRUE ~ SPCD))


## M. tripetala SPCD == 658

matr.plots <- fia$TREE %>% 
  filter(SPCD == 658, most.recent=="yes") %>% 
  pull(PLT_CN) %>% 
  unique()

```

Loading some spatial data for plotting
```{r}

# WGS84
old.proj <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"

# Albers Equal Area; centered in western US
base.proj <- "+proj=aea +lat_0=23 +lon_0=-96 +lat_1=29.5 +lat_2=45.5 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs"

# US state boundaries
states <- readOGR(dsn="/Users/DanielPerret/Box/01. daniel.perret Workspace/base_spatialdata/state_boundaries",
                  layer = "state_boundaries", verbose=F) 

# %>% 
#   spTransform(.,
#               CRSobj = CRS(base.proj))

# North American continent
cont <- readOGR(dsn="/Users/DanielPerret/Box/01. daniel.perret Workspace/base_spatialdata/continents",
                layer = "na",
                verbose=F,
                p4s = old.proj)

# %>% 
#   spTransform(.,
#               CRSobj = CRS(base.proj))


range <- sf::st_read("/Users/DanielPerret/Box/01. daniel.perret Workspace/base_spatialdata/named_ranges/magntrip.shp", crs = old.proj)

```




Making some basic summaries requested by Dov:

"FIA analysis of plots in the coastal plains (eastern 1/3) of VA, NC and or SC for Magnolia tripetala; if it turns out there are not enough coastal plans plots to examine then we could look at the Piedmont (roughly center third of these states) instead; and worst case scenario (but still not a bad scenario) is we look at plots in the mountains.
    
    a. mean (and full range, i.e. max and min observed at any plot) of the basal area and or number of adult/sapling trees on a subplot when they are present
    
    b. (not as important) mean number of MT seedlings found on subplots that have MT"

```{r}

t.est <- rFIA::tpa(db = fia,
                   treeDomain = SPCD==658,
                   #DIA > 5,
                   byPlot = T,
                   totals=T) %>% 
  filter(TPA > 0) %>% 
  group_by(pltID) %>% 
  filter(YEAR == max(YEAR)) %>% 
  left_join(fia$PLOT %>% 
              select(PLT_CN, state_key)) %>% 
  mutate(region = case_when(state_key %in% c("VA","NC") ~ "Piedmont",
                            state_key %in% c("WV", "KY","TN") ~ "Mountains",
                            state_key %in% c("GA","AL") ~ "South"))

```

Quick map
```{r}

matr.sp <- t.est %>% 
  left_join(fia$PLOT %>% select(PLT_CN,LON,LAT),
            by="PLT_CN") %>% 
  SpatialPointsDataFrame(coords = .[,c("LON","LAT")],data = ., proj4string = CRS(old.proj)) %>% 
  sf::st_as_sf()

ggplot() +
  geom_sf(data = cont %>% # shapefile of north american continent
            as(.,"sf"),
          col=linecolor,
          fill = mapcolor,
          lwd=0.3) +  
  geom_sf(data = states %>% # shapefile of US state boundaries
            as(.,"sf"),
          fill=NA,
          col=linecolor) +
  geom_sf(data = range,
          fill = "gray85",
          col=NA)+
  geom_sf(data = matr.sp,
          aes(bg = BAA),
          pch = 21,
          alpha = 0.5,
          size = 4)+
  lims(x = c(-90, -75),
       y = c(30,40)) +
  scale_color_binned(type = "viridis",aesthetics="bg",breaks=c(0,5,10,15,20,25,30)) +
  theme(legend.position="right",
        panel.background = element_rect(fill="skyblue1"))  

ggplot() +
  geom_sf(data = cont %>% # shapefile of north american continent
            as(.,"sf"),
          col=linecolor,
          fill = mapcolor,
          lwd=0.3) +  
  geom_sf(data = states %>% # shapefile of US state boundaries
            as(.,"sf"),
          fill=NA,
          col=linecolor) +
  geom_sf(data = range,
          fill = "gray85",
          col=NA)+
  geom_sf(data = matr.sp,
          aes(bg = region),
          pch = 21,
          alpha = 0.5,
          size = 4)+
  lims(x = c(-90, -75),
       y = c(30,40)) +
  theme(legend.position="right",
        panel.background = element_rect(fill="skyblue1"))  


```

Quick density plots for Dov

```{r}

t.est <- t.est %>% 
  left_join(fia$PLOT %>% 
              select(PLT_CN, state_key)) %>% 
  mutate(region = case_when(state_key %in% c("VA","NC") ~ "Piedmont",
                            state_key %in% c("WV", "KY","TN") ~ "Mountains",
                            state_key %in% c("GA","AL") ~ "South"))

t.est %>% 
  sf::st_drop_geometry() %>% 
  ggplot(.,
         aes(x = BAA,
             groups=region,
             fill = region)) +
  geom_density(col = "gray35",
               alpha = 0.5)

t.est %>% 
  sf::st_drop_geometry() %>% 
  ggplot(.,
         aes(x = TPA,
             groups=region,
             fill = region)) +
  geom_density(col = "gray35",
               alpha = 0.5)

fia$TREE %>% 
  filter(PLT_CN %in% t.est$PLT_CN,
         SPCD == 658,
         STATUSCD == 1) %>% 
    mutate(region = case_when(state_key %in% c("VA","NC") ~ "Piedmont",
                            state_key %in% c("WV", "KY","TN") ~ "Mountains",
                            state_key %in% c("GA","AL") ~ "South")) %>% 
  ggplot(.,
         aes(x = DIA,
             groups = region,
             fill = region)) +
  geom_density(col = "gray35",
               alpha = 0.5)

```

Summaries

```{r}

t.est %>% 
  group_by(region) %>% 
  summarise(meanBA = mean(BAA),
            minBA = min(BAA),
            maxBA = max(BAA),
            sdBA = sd(BAA),
            meanTPA = mean(TPA),
            minTPA = min(TPA),
            maxTPA = max(TPA),
            sdTPA = sd(TPA)) %>% view()

fia$TREE %>% 
  filter(PLT_CN %in% t.est$PLT_CN,
         SPCD == 658,
         STATUSCD == 1) %>% 
    mutate(region = case_when(state_key %in% c("VA","NC") ~ "Piedmont",
                            state_key %in% c("WV", "KY","TN") ~ "Mountains",
                            state_key %in% c("GA","AL") ~ "South")) %>% 
  group_by(region) %>% 
  summarise(meanDIA = mean(DIA),
            sdDIA = sd(DIA),
            minDIA = min(DIA),
            maxDIA = max(DIA)) %>% view()
```

















